\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces}

### 2025-06-18 paper/manuscript.qmd
add paragraph on metacognitive stuff

However, one might be concerned about the application of the CHOCO model (a mixture of two continuous Beta distributions) to ordinal data in Sample 2 (where a 7-point Likert scale was used). Ordinal ratings (limited number of discrete options represented numerically by integers) are commonly used in psychology, and often modeled as continuous. While this fact has spurred its own polarized debate [@norman2010likert; @sullivan2013analyzing; @liddell2018analyzing], our position is that the model used should ideally reflect the data *generating* mechanism rather than necessarily focus on the *observed* data^[Ideally, the data recording method should be aligned with the assumed generating mechanism, which is an experiment design issue rather than a statistical one.] - although the former cannot always be easily inferred or known. 
For the data of Sample 2, we held the assumption that the underlying latent distribution was CHOCO distributed (as there is no reason to assume it would be different from Sample 1), and thus treated as continuous CHOCO-distributed data, despite the fact that it was collected on a 7-point scale. 
The motivation was to see if the model could extrapolate and reliably estimate all the parameters based on a limited number of data points but means in our case that, effectively, each Beta distribution was extrapolated from the frequency of only three unique values. 
While this is far from ideal, the CHOCO model still proved insightful, although it might have impacted its ability detect finer-grain changes, including quadratic relationships.

@@

First, future studies should investigate the psychometric quality of the CHOCO model to derive practical guidelines for its optimal use. These include assessing the minimum amount of data required to reliably estimate and recover the different CHOCO parameters. When applying the model to ordinal data (such as Likert scales), another key question is the minimum number of unique response categories needed to detect subtle or nonlinear effects, such as quadratic modulation of confidence. Second, assessing the model’s robustness when applied to unimodal or skewed data would extend its utility beyond strictly bimodal contexts and support its general applicability. Third, more parsimonious variants of the model could be explored, such as versions with a single shared precision parameter for both sides or formulations where the confidence on one side is modeled as a function of the other. Such simplifications may retain interpretability while reducing parameter overhead and improving model stability.

In conclusion, we developed and validated the CHOCO model, a theoretically grounded and practically flexible tool for modeling subjective judgments and delineating choice from confidence processes. It not only advances statistical modeling of decision-making data but opens new avenues for dissecting the cognitive mechanisms of belief, confidence, and reality perception.

>>>>

However, one might be concerned about the application of the CHOCO model (a mixture of two continuous Beta distributions) to ordinal data in Sample 2 (where a 7-point Likert scale was used). Ordinal ratings (limited number of discrete options represented numerically by integers) are commonly used in psychology, and often modeled as continuous. While this fact has spurred its own polarized debate [@norman2010likert; @sullivan2013analyzing; @liddell2018analyzing; @rausch2014comparison], our position is that the model used should ideally reflect the data *generating* mechanism rather than necessarily focus on the *observed* data^[Ideally, the data recording method should be aligned with the assumed generating mechanism, which is an experiment design issue rather than a statistical one.] - although the former cannot always be easily inferred or known. 
For the data of Sample 2, we held the assumption that the underlying latent distribution was CHOCO distributed (as there is no reason to assume it would be different from Sample 1), and thus treated as continuous CHOCO-distributed data, despite the fact that it was collected on a 7-point scale. 
The motivation was to see if the model could extrapolate and reliably estimate all the parameters based on a limited number of data points but means in our case that, effectively, each Beta distribution was extrapolated from the frequency of only three unique values. 
While this is far from ideal, the CHOCO model still proved insightful, although it might have impacted its ability detect finer-grain changes, including quadratic relationships.

@@

Firstly, future studies should investigate the psychometric quality of the CHOCO model to derive practical guidelines for its optimal use. These include assessing the minimum amount of data required to reliably estimate and recover the different CHOCO parameters. When applying the model to ordinal data (such as Likert scales), another key question is the minimum number of unique response categories needed to detect subtle or nonlinear effects, such as quadratic modulation of confidence. Second, assessing the model’s robustness when applied to unimodal or skewed data would extend its utility beyond strictly bimodal contexts and support its general applicability. Third, more parsimonious variants of the model could be explored, such as versions with a single shared precision parameter for both sides or formulations where the confidence on one side is modeled as a function of the other. Such simplifications may retain interpretability while reducing parameter overhead and improving model stability.

In the domain of reaction times, researchers distinguish between "descriptive" models (e.g., ex-Gaussian or log-normal distributions) that capture characteristic distributional shapes, and "data-generating" models like the Drift Diffusion Model [DDM, @ratcliff2016diffusion] or Linear Ballistic Accumulator [LBA, @brown2008simplest], which simulate an assumed generative process (typically, sequential evidence accumulation). Drawing a comparison with this distinction, CHOCO was designed as a descriptive model to flexibly fit bimodal confidence data from subjective scales where responses implicitly reflect two categories. However, CHOCO remains agnostic about the underlying cognitive dynamics that generate those confidence ratings. In contrast, the field of metacognition has developed cognitive models that conceptualize confidence generation as a second-order inference. These include approaches derived from Signal Detection Theory, such as the meta-d' method which quantifies metacognitive sensitivity by assessing how well confidence discriminates between correct and incorrect decisions [@maniscalco2012signal; @fleming2017hmeta], as well as dynamic models like the Weighted Evidence and Visibility model or the Post-Decisional Accumulation model, which propose that confidence arises from mechanisms like post-decisional evidence sampling [@rausch2021modelling]. A critical distinction is that these models are typically applied to two-step paradigms, where participants first make a categorical decision and then separately report their confidence. Future studies could investigate the differences and potential overlap in the applicability area and insights provided by these two classes of models.

In conclusion, we developed and validated the CHOCO model, a mathematically grounded and practically flexible tool for modeling subjective judgments and delineating choice from confidence processes. It not only advances statistical modeling of decision-making data but opens new avenues for dissecting the cognitive mechanisms of belief, confidence, and reality perception.

### 2025-06-18 paper/manuscript.tex
add paragraph on metacognitive stuff

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, breakable, arc=.35mm, colframe=quarto-callout-note-color-frame, colback=white, leftrule=.75mm, opacityback=0, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm]

@@

However, one might be concerned about the application of the CHOCO model
(a mixture of two continuous Beta distributions) to ordinal data in
Sample 2 (where a 7-point Likert scale was used). Ordinal ratings
(limited number of discrete options represented numerically by integers)
are commonly used in psychology, and often modeled as continuous. While
this fact has spurred its own polarized debate
(\citeproc{ref-liddell2018analyzing}{Liddell \& Kruschke, 2018};
\citeproc{ref-norman2010likert}{Norman, 2010};
\citeproc{ref-sullivan2013analyzing}{Sullivan \& Artino Jr, 2013}), our
position is that the model used should ideally reflect the data
\emph{generating} mechanism rather than necessarily focus on the
\emph{observed} data\footnote{Ideally, the data recording method should
  be aligned with the assumed generating mechanism, which is an
  experiment design issue rather than a statistical one.} - although the
former cannot always be easily inferred or known. For the data of Sample
2, we held the assumption that the underlying latent distribution was
CHOCO distributed (as there is no reason to assume it would be different
from Sample 1), and thus treated as continuous CHOCO-distributed data,
despite the fact that it was collected on a 7-point scale. The
motivation was to see if the model could extrapolate and reliably
estimate all the parameters based on a limited number of data points but
means in our case that, effectively, each Beta distribution was
extrapolated from the frequency of only three unique values. While this
is far from ideal, the CHOCO model still proved insightful, although it
might have impacted its ability detect finer-grain changes, including
quadratic relationships.

@@

First, future studies should investigate the psychometric quality of the
CHOCO model to derive practical guidelines for its optimal use. These
include assessing the minimum amount of data required to reliably
estimate and recover the different CHOCO parameters. When applying the
model to ordinal data (such as Likert scales), another key question is
the minimum number of unique response categories needed to detect subtle
or nonlinear effects, such as quadratic modulation of confidence.
Second, assessing the model's robustness when applied to unimodal or
skewed data would extend its utility beyond strictly bimodal contexts
and support its general applicability. Third, more parsimonious variants
of the model could be explored, such as versions with a single shared
precision parameter for both sides or formulations where the confidence
on one side is modeled as a function of the other. Such simplifications
may retain interpretability while reducing parameter overhead and
improving model stability.

In conclusion, we developed and validated the CHOCO model, a
theoretically grounded and practically flexible tool for modeling
subjective judgments and delineating choice from confidence processes.
It not only advances statistical modeling of decision-making data but
opens new avenues for dissecting the cognitive mechanisms of belief,
confidence, and reality perception.

@@

@@

@@

@@

>>>>

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, breakable, opacityback=0, colframe=quarto-callout-note-color-frame, arc=.35mm, colback=white, left=2mm, rightrule=.15mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm]

@@

However, one might be concerned about the application of the CHOCO model
(a mixture of two continuous Beta distributions) to ordinal data in
Sample 2 (where a 7-point Likert scale was used). Ordinal ratings
(limited number of discrete options represented numerically by integers)
are commonly used in psychology, and often modeled as continuous. While
this fact has spurred its own polarized debate
(\citeproc{ref-liddell2018analyzing}{Liddell \& Kruschke, 2018};
\citeproc{ref-norman2010likert}{Norman, 2010};
\citeproc{ref-rausch2014comparison}{Rausch \& Zehetleitner, 2014};
\citeproc{ref-sullivan2013analyzing}{Sullivan \& Artino Jr, 2013}), our
position is that the model used should ideally reflect the data
\emph{generating} mechanism rather than necessarily focus on the
\emph{observed} data\footnote{Ideally, the data recording method should
  be aligned with the assumed generating mechanism, which is an
  experiment design issue rather than a statistical one.} - although the
former cannot always be easily inferred or known. For the data of Sample
2, we held the assumption that the underlying latent distribution was
CHOCO distributed (as there is no reason to assume it would be different
from Sample 1), and thus treated as continuous CHOCO-distributed data,
despite the fact that it was collected on a 7-point scale. The
motivation was to see if the model could extrapolate and reliably
estimate all the parameters based on a limited number of data points but
means in our case that, effectively, each Beta distribution was
extrapolated from the frequency of only three unique values. While this
is far from ideal, the CHOCO model still proved insightful, although it
might have impacted its ability detect finer-grain changes, including
quadratic relationships.

@@

Firstly, future studies should investigate the psychometric quality of
the CHOCO model to derive practical guidelines for its optimal use.
These include assessing the minimum amount of data required to reliably
estimate and recover the different CHOCO parameters. When applying the
model to ordinal data (such as Likert scales), another key question is
the minimum number of unique response categories needed to detect subtle
or nonlinear effects, such as quadratic modulation of confidence.
Second, assessing the model's robustness when applied to unimodal or
skewed data would extend its utility beyond strictly bimodal contexts
and support its general applicability. Third, more parsimonious variants
of the model could be explored, such as versions with a single shared
precision parameter for both sides or formulations where the confidence
on one side is modeled as a function of the other. Such simplifications
may retain interpretability while reducing parameter overhead and
improving model stability.

In the domain of reaction times, researchers distinguish between
``descriptive'' models (e.g., ex-Gaussian or log-normal distributions)
that capture characteristic distributional shapes, and
``data-generating'' models like the Drift Diffusion Model (DDM,
\citeproc{ref-ratcliff2016diffusion}{Ratcliff et al., 2016}) or Linear
Ballistic Accumulator (LBA, \citeproc{ref-brown2008simplest}{Brown \&
Heathcote, 2008}), which simulate an assumed generative process
(typically, sequential evidence accumulation). Drawing a comparison with
this distinction, CHOCO was designed as a descriptive model to flexibly
fit bimodal confidence data from subjective scales where responses
implicitly reflect two categories. However, CHOCO remains agnostic about
the underlying cognitive dynamics that generate those confidence
ratings. In contrast, the field of metacognition has developed cognitive
models that conceptualize confidence generation as a second-order
inference. These include approaches derived from Signal Detection
Theory, such as the meta-d' method which quantifies metacognitive
sensitivity by assessing how well confidence discriminates between
correct and incorrect decisions
(\citeproc{ref-fleming2017hmeta}{Fleming, 2017};
\citeproc{ref-maniscalco2012signal}{Maniscalco \& Lau, 2012}), as well
as dynamic models like the Weighted Evidence and Visibility model or the
Post-Decisional Accumulation model, which propose that confidence arises
from mechanisms like post-decisional evidence sampling
(\citeproc{ref-rausch2021modelling}{Rausch et al., 2021}). A critical
distinction is that these models are typically applied to two-step
paradigms, where participants first make a categorical decision and then
separately report their confidence. Future studies could investigate the
differences and potential overlap in the applicability area and insights
provided by these two classes of models.

In conclusion, we developed and validated the CHOCO model, a
mathematically grounded and practically flexible tool for modeling
subjective judgments and delineating choice from confidence processes.
It not only advances statistical modeling of decision-making data but
opens new avenues for dissecting the cognitive mechanisms of belief,
confidence, and reality perception.

@@

\bibitem[\citeproctext]{ref-brown2008simplest}
Brown, S. D., \& Heathcote, A. (2008). The simplest complete model of
choice response time: Linear ballistic accumulation. \emph{Cognitive
Psychology}, \emph{57}(3), 153--178.

@@

\bibitem[\citeproctext]{ref-fleming2017hmeta}
Fleming, S. M. (2017). HMeta-d: Hierarchical bayesian estimation of
metacognitive efficiency from confidence ratings. \emph{Neuroscience of
Consciousness}, \emph{2017}(1), nix007.

@@

\bibitem[\citeproctext]{ref-maniscalco2012signal}
Maniscalco, B., \& Lau, H. (2012). A signal detection theoretic approach
for estimating metacognitive sensitivity from confidence ratings.
\emph{Consciousness and Cognition}, \emph{21}(1), 422--430.

@@

\bibitem[\citeproctext]{ref-ratcliff2016diffusion}
Ratcliff, R., Smith, P. L., Brown, S. D., \& McKoon, G. (2016).
Diffusion decision model: Current issues and history. \emph{Trends in
Cognitive Sciences}, \emph{20}(4), 260--281.

\bibitem[\citeproctext]{ref-rausch2021modelling}
Rausch, M., Hellmann, S., \& Zehetleitner, M. (2021). Modelling
visibility judgments using models of decision confidence.
\emph{Attention, Perception, \& Psychophysics}, \emph{83}(8),
3311--3336.

\bibitem[\citeproctext]{ref-rausch2014comparison}
Rausch, M., \& Zehetleitner, M. (2014). A comparison between a visual
analogue scale and a four point scale as measures of conscious
experience of motion. \emph{Consciousness and Cognition}, \emph{28},
126--140.
